---
author: "Afrah Shafquat"
output: html_document
---

Quantitative Genomics and Genetics 2017
======

Computer Lab 4
------

-- 23 February 2017

-- Author: Afrah Shafquat (as3397@cornell.edu)


### 1. Boolean Data Type

- Boolean data type is a data type with only two possible values, TRUE and FALSE.

- Its main usage is for testing conditions. 

- The data type is known as logicals in R

- In R and many other languages, the numeric values for TRUE and FALSE are 1 and 0.

- T and TRUE are the same, as is FALSE and F (Unless you define F and/or T as variable names. Then its the variable you defined. If you dont understand this, come talk to me)


- This can be illustrated as follows

```{r, comment = NA}

sum(TRUE + TRUE)

as.logical(1)

sum(FALSE)

as.numeric(TRUE)

class(TRUE)

T==TRUE

```

- Booleans are useful in testing conditions. 

- For example, if you are interested in finding out if certain elements of a vector are greater than or smaller than a certain value, you can use >, <, >=, <=

```{r, comment = NA}
example.vector <- seq(1,25,by= 2)
example.vector
example.vector > 10
example.vector >= 15
example.vector <= 5
```

- Each position where the value meets the condition will be marked with TRUE, and with FALSE otherwise.

- Since TRUE and FALSE are essentially 1 and 0, you can easily find out how many elements satisfy the condition by simply taking the sum of the result. 

```{r, comment = NA}

sum(example.vector >10)

```

- The same applies to matrices

```{r, comment = NA}
example.mx <- matrix(c(2,5,7,-2,-5,-10), ncol = 3, byrow=T)
example.mx < 1
```

- You can also check if an element is equal to a specific value 

```{r, comment = NA}
dim(example.mx)[1] == 2
```

```{r, comment = NA}
example.vector == 3
```

### 2. Boolean Algebra

- Boolean algebra allows you to combine multiple conditions.

- There are three basic operations AND (&), OR (|) and NOT (!).

- The AND (&) operator returns TRUE **only if all conditions are TRUE**

```{r, comment = NA}
FALSE & FALSE
TRUE & FALSE
TRUE & TRUE

# Example of an AND operator
example.vector > 5 & example.vector < 10

# if you want to see the actual elements 
example.vector[example.vector >10 & example.vector < 20]

```

- The OR (|) operator returns TRUE **when at least one condition is TRUE**

```{r, comment= NA}

FALSE | FALSE
TRUE | FALSE
TRUE | TRUE

# Example of an OR operator
example.vector > 10 | example.vector < 20

example.vector < 10 | example.vector > 20

```

- NOT (!) returns the opposite result

```{r, comment = NA}
!(TRUE)

example.vector != 3
```

### 3. More on Vector Elements

- If you want to check whether a certain element is present or absent in a vector use the %in% operator

```{r, comment = NA}

fruits <- c("banana","apple","strawberry","peach","mango")

"mango" %in% fruits

"durian" %in% fruits
```

- We can see what the ! operator is doing by wrapping the previous expression with !() 

```{r, comment = NA}
!("durian" %in% fruits)

```

- You can find out the index of a certain entry in a vector by using the which() function 

```{r, comment = NA}
which(fruits == "apple")
```

- If you want to compare two vectors,

```{r, comment = NA}
fruits2 <- c("orange","banana","durian","cherry","mango","apple", 'apple')

fruits %in% fruits2

fruits2 %in% fruits

# show me the position
which(fruits2 %in% fruits)
which(fruits %in% fruits2)

#show me the elements
fruits2[fruits2 %in% fruits]

# There is also a function for this
#note that apple doesnt show up twice! #settheoryisawesome
intersect(fruits2, fruits)

```

More on functions for set operations:

```{r}
?intersect
```

### 2. If / else statements 

- By using if and else statements you can insert condition specific executions in your script

- The code inside an if statement will only be executed when the condition is TRUE

- The structure looks like this

```{r, comment = NA, eval = FALSE}
if (condition) {
  do stuff
} else {
  do stuff
}

# OR you can add more levels by using else if

if(condition){
  do stuff
} else if (condition 2){
  do stuf
} else {
  do stuff
}
```


- Here is a simple example

```{r, comment = NA}

# Loop over individual elements in example.vector
for( i in example.vector){
  
    if( i < 10 ){
        cat(i, "is smaller than 10 \n") 
    } else if ( 10 <= i & i < 20){
        cat(i, "is in the interval [10,20) \n")
    } else {
        cat(i, "is larger than 20 \n")
    }

}


```


### 4. More on plotting (if time permits)

- Library ggplot2 allows you to plot figures using dataframes. You can create the same plots using functions: plot and other functions. However, ggplot allows more intuitive control and has aesthetically pleasing defaults.

- To install the package, use install.packages('ggplot2', dependencies=T)

- Inputs are dataframes.

- To plot a scatterplot

```{r, comment = NA, fig.align='center',fig.width = 4,fig.height = 4}

library(ggplot2)
x=rnorm(1000,0,100)
y=rnorm(1000,5,20)

df1 = data.frame(x=x,y=y)

ggplot(data=df1)+geom_point(aes(x=x,y=y))+labs(x='xlabel',y='ylabel', title='Scatter plot')

```

- To plot a line

```{r, comment = NA, fig.align='center',fig.width = 4,fig.height = 4}

ggplot(data=df1)+geom_line(aes(x=x,y=y))+labs(x='xlabel',y='ylabel', title='Line plot')

```

- To plot a histogram

```{r, comment = NA, fig.align='center',fig.width = 4,fig.height = 4}

ggplot(data=df1)+geom_histogram(aes(x=x))+labs(x='xlabel',y='ylabel', title='Histogram')

```

- You can do numerous customizations: size, colors etc.

```{r, comment = NA, fig.align='center',fig.width = 7,fig.height = 4}

ggplot(data=df1)+geom_point(aes(x=x,y=y, size=x, color=x))+labs(x='xlabel',y='ylabel', title='Scatter plot')

```

- As you can use par(mfrow=c(rows, columns)) for plot to provide a grid or matrix for plotting your figures. The parallel for ggplot is grid.

- You can put your ggplot object/plot into a variable. This variable is now a grob or a graphical object. You can "arrange" your grobs into a lattice or a grid using the function grid.arrange


```{r, comment = NA, fig.align='center',fig.width = 10,fig.height = 10}

library(gridExtra)
g4<-ggplot(data=df1)+geom_point(aes(x=x,y=y, size=x, color=x))+labs(x='xlabel',y='ylabel', title='Scatter plot')
g3<-ggplot(data=df1)+geom_histogram(aes(x=x))+labs(x='xlabel',y='ylabel', title='Histogram')
g2<-ggplot(data=df1)+geom_line(aes(x=x,y=y))+labs(x='xlabel',y='ylabel', title='Line plot')
g1<-ggplot(data=df1)+geom_point(aes(x=x,y=y))+labs(x='xlabel',y='ylabel', title='Scatter plot')

grid.arrange(g1,g2,g3,g4, nrow=2)

```


- For more information on ggplot2, refer to the documentation (http://docs.ggplot2.org/current)

### Exercise

- Using if/else statements and plotting functions create a function that can generate a histogram, a scatter plot, or a density plot depending on the input. 


```{r, include= FALSE,comment = NA}

data_plotter <- function(input_data, plot_type = "histogram"){
  
    if(plot_type == "histogram"){
        hist(input_data, main = "Histogram of input data")  
    } else if (plot_type == "scatter"){
        plot(input_data, type = 'p', main = "scatter plot of input data")
    } else if (plot_type == "density"){
        plot(density(input_data), main = "Density plot of input data", 
             type = 'l', 
             xlim = c(min(input_data), 
                      max(input_data)))
    }
}

```

```{r, fig.align='center', fig.height = 4, fig.width = 4}

sample_data <- runif(1000)

data_plotter(sample_data, plot_type = "histogram")
data_plotter(sample_data, plot_type = "scatter")
data_plotter(sample_data, plot_type = "density")

```

